Article Title

### 2014-11-21 manuscript.tex
fixed error in equation 17
Ready for second resubmission

%% Include all macros below. Please limit the use of macros.
\DeclareMathOperator{\Tr}{tr}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)}
\newcommand{\T}{{\sf T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\DeclareMathOperator*{\argmin}{arg\,min}
%% END MACROS SECTION

@@

The evaluation of any covariance matrix estimator, $C$, is performed with respect to a \emph{loss function} $\loss{C,\Sigma}$ to quantify its discrepancy from truth, $\Sigma$.  The loss function must attain its minimum when $C=\Sigma$.

In this study, we adopted the \emph{normal loss} function defined as
\begin{equation}\label{eq:loss}
    \loss{C,\Sigma} = \frac 1 p\left[\ln \det C + \Tr(C^{-1}\Sigma)\right]
\end{equation}

This loss function is related to the multivariate normal log-likelihood function $L(\Sigma|C_{\sf sample})$,
\begin{equation}
    \loss{X,Y} \equiv -\ln(2\pi) - \frac 2 {pn} L(X|Y)
\end{equation}
Normalization by $p$ and $n$ makes the values of the loss function comparable across different population sizes and sample sizes.

Although the two functions are similar in form, they differ conceptually: The log-likelihood $L(\Sigma|C_{\sf sample})$ is a function of the parameter $\Sigma$ given the sample covariance matrix $C_{\sf sample}$ whereas the loss function $\loss{C,\Sigma}$ expresses the deviation of an arbitrary estimate $C$ from the truth $\Sigma$. Minimum expected loss may be found at a point that is far removed from the point of maximum likelihood given the same data.

The choice of the normal loss function is motivated, in part, by mathematical convenience. We expect that the main conclusions of our study will not change qualitatively with other well behaved loss functions, such as Stein's entropy loss or quadratic loss \cite{James:1961, Fan:2008, Ledoit:2004, Schafer:2005}.

We drew 30 independent samples of sizes $n=250$, 500, 1000, 2000, and 4000 from each model and computed the \emph{excess loss} $\loss{C,\Sigma}-\loss{\Sigma,\Sigma}$ for each of the five estimators.  The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  All the regularized estimators produced better estimates (lower excess loss) than the sample covariance matrix.  However, estimators whose structure matched the true model outperformed the other estimators (Fig.~\ref{fig:1} Row 5).  Note that when the ground truth had zero correlations (Column A), $C_{\sf factor}$ performed equally well to $C_{\sf diag}$ because it correctly identified zero factors and only estimated the individual variances. Similarly, when the number of latent units was zero (Column C), $C_{\sf sparse+latent}$ performed equally well to $C_{\sf sparse}$ because it correctly identified zero latent units.

With increasing sample sizes, all estimators converged to the truth (zero excess loss).  However,
the discriminability between their performances increased with sample size (data not shown).

With empirical data, the ground truth, $\Sigma$, is not accessible and excess loss cannot be computed directly. Instead, the loss function may be estimated from the sample through \emph{validation}.  In a validation procedure, an independent \emph{testing sample} is set aside to compute an additional sample covariance estimate $C_{\sf sample}^\prime$ to validate the estimate $C$ computed from the  \emph{training sample}.   The estimate $C_{\sf sample}^\prime$ is used to compute the \emph{validation loss} $\loss{C,C_{\sf sample}^\prime}$, which approximates the loss  $\loss{C,\Sigma}$.

The normal loss (Eq.~\ref{eq:loss}) is particularly suitable because it is additive in its second argument, \emph{i.e.} 
\begin{equation*}\label{eq:additivity}
    \loss{C,Z_1} + \loss{C,Z_2} \equiv \loss{C,Z_1+Z_2}
\end{equation*}
Then, with fixed $C$, validation loss is an unbiased estimate of true loss:
\begin{equation*}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses; their validation losses are biased estimates of the loss function and validation does not exactly evaluate a discrepancy from truth.

The loss function above assumes a single covariance matrix across all conditions in the dataset.  However, neuroscientists often estimate a common correlation matrix across multiple stimulus conditions when the variances of responses are conditioned on the stimulus \cite{Vogels:1989, Ponce:2013}. In this case, the stimulus-dependent means and variances are estimated from the training dataset along with a common correlation matrix. An adaptation of the normal loss function to this case is described in Methods.

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the testing sample with the other $K-1$ subsets serving as the training sample. The validation losses from all such `folds' are averaged to produce the \emph{cross-validation loss}.

Cross-validation accurately reproduced relative values of the loss and the rankings of estimators without access to the ground truth (Fig.~\ref{fig:1} Row 6).

To demonstrate that estimator rankings did not strongly depend on the gaussianity of the data-generating process, we repeated the cross-validated evaluation after replacing the Gaussian data-generating models with pairwise Ising models. 
Ising models have been used to infer functional connectivity from neuronal spike trains \cite{Hertz:2011}. 
Conveniently, the Ising model has equivalent mathematical form to the Gaussian distribution,
\begin{equation}\label{eq:ising}
    x\sim \frac 1 {Z(S,h)} \exp\left(-\frac 1 2 x^\T S x + h^\T x \right)
\end{equation}
but the Ising model is defined on the multivariate binary domain $x\in\{-1,+1\}^p$ rather than the continuous domain of the Gaussian  model. 
Both models are maximum-entropy models constrained by the mean firing rates and the covariance matrix \cite{Jaynes:1957}.
The partition function $Z(S,h)$ normalizes the distribution on the models' respective domains. 
In the Gaussian model, the matrix $S$ is the precision matrix or the inverse of the covariance matrix; and the mean values are $\mu=-S^{-1}h$.  
For the Ising model, $-S$ is the matrix of pairwise interactions and $h$ is the vector of the cells' individual activity drives but they do not have a simple relationship to the means and the covariance matrix. 
Both distributions have the same structure of pairwise conditional dependencies: zeros in the matrix $S$ indicate conditional independence between the corresponding pair of neurons. 
Despite the departure from strictly linear dependencies between the variables characteristic of the Gaussian distribution, the Ising model yielded the same rankings of estimators as the Gaussian model in cross-validation (Fig.~S1).

This simulation study demonstrated that, with sufficiently large sample sizes, imposing the correct type of structure on the estimate leads to greater estimation improvement than imposing other structures.

\subsection*{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (Fig.~\ref{fig:2} A--B) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (Fig.~\ref{fig:2} C and D).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.~\ref{fig:2} C). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (Fig.~\ref{fig:2} D). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (Fig.~\ref{fig:2} E and F). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (Fig.~\ref{fig:5} D).

@@

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Fig.~S2 and  Methods). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<0.01$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

\subsection*{Structure of $C_{\sf sparse+latent}$ estimates}
We examined the composition of the $C_{\sf sparse+latent}$ estimates at each imaged site (Fig.~\ref{fig:4} and Fig.~\ref{fig:5}). Although the regularized estimates were similar to the sample correlation matrix (Fig.~\ref{fig:4} A and B), the corresponding partial correlation matrices differed substantially (Fig.~\ref{fig:4} C and D). The estimates separated two sources of correlations: a network of linear interactions expressed by the sparse component of the inverse and latent units expressed by the low-rank components of the inverse (Fig.~\ref{fig:4} E). The sparse partial correlations revealed a network that differed substantially from the network composed of the greatest coefficients in the sample correlation matrix (Fig.~\ref{fig:4} F, G, H, and I).

In the example site (Fig.~\ref{fig:4}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $\mbox{connectivity}=1-\mbox{sparsity}$) with average node degree of 20.9 (Fig.~\ref{fig:4} G). The average node degree, \emph{i.e.}\;the average number of interactions linking each neuron, is related to connectivity as $\mbox{degree} = \mbox{connectivity}\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units increased with population size (Fig.~\ref{fig:5} A) but the connectivity was highly variable (Fig.~\ref{fig:5} B): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and warrants further investigation.

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 27 sites were about 5 times lower than the average sample correlations (Fig.~\ref{fig:5} C). This suggests that correlations between neurons build up from multiple chains of smaller interactions. Furthermore, the average partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.45 whereas that of the average partial correlations was 0.29, with larger populations exhibiting greater uniformity of average partial correlations than the smaller populations.

While the sample correlations were mostly positive, the sparse component of the partial correlations (`interactions') had a high fraction (28.7\% in the example site) of negative values (Fig.~\ref{fig:4} F). The fraction of negative interactions increased with the inferred connectivity (Fig.~\ref{fig:5} D), suggesting that negative interactions can be inferred only after a sufficient density of positive interactions has been uncovered.

Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013, Sadovsky:2014}.  We therefore compared the interactions in the sparse component of $C_{\sf sparse+latent}$ to those obtained from the sample correlations thresholded to the same level of connectivity. The networks revealed by the two methods differed substantially. In the example site with 7.2\% connectivity in $C_{\sf sparse+latent}$, only 27.7\% of the connections coincided with the above-threshold sample correlations (Fig.~\ref{fig:4} F, H, and I). In particular, most of the inferred negative interactions corresponded to low sample correlations (Fig.~\ref{fig:4} F) where high correlations should be expected given the rest of the correlation matrix.

@@

We examined how the structure of the $C_{\sf sparse+latent}$ estimates related to the differences in orientation preference and to the physical distances separating pairs of cells (Fig.\;\ref{fig:6}).  Five sites with highest pairwise connectivities were included in the analysis. Partial correlations were computed using Eq.~\ref{eq:partial} based on the regularized estimate, including both the sparse and the latent component. Connectivity was computed as the fraction of pairs of cells connected by non-zero elements (interactions) in the sparse component of the estimate, distinguishing between the positive and negative connectivities.

First, we analyzed how correlations and connectivity depended on the difference in preferred orientations ($\Delta \mbox{ori}$) of pairs of significantly ($\alpha=0.05$) tuned cells. The partial correlations decayed more rapidly with $\Delta\mbox{ori}$ than did sample correlations ($p<10^{-9}$ in each of the five sites, two-sample $t$-test of the difference of the linear regression coefficients). Positive connectivity decreased with $\Delta\mbox{ori}$ ($p<0.005$ in each of the five sites, $t$-test on the logistic regression coefficient) whereas negative connectivity did not decrease (Fig.~\ref{fig:6} D): The slope in the logistic model of connectivity with respect to $\Delta\mbox{ori}$ was significantly higher for positive than for negative interactions ($p<0.04$ in each of the five sites, two-sample $t$-test of the difference of the logistic regression coefficient).

Second, we compared how correlations and connectivity depended on the physical distance separating pairs of cells. We distinguished between lateral distance, $\Delta x$, in the plane parallel to the pia, and vertical distance, $\Delta z$, orthogonal to the pia.  When considering the dependence on $\Delta x$, the analysis was limited to cell pairs located at the same depth with $\Delta z < 30\,\mu\mbox{m}$; conversely, when considering the dependence on $\Delta z$, only vertically aligned cell pairs with $\Delta x < 30\,\mu\mbox{m}$ were included. Again, the partial correlations decayed more rapidly both laterally and vertically than sample correlations ($p<10^{-6}$ in each of the five sites, for both lateral and vertical distances, two-sample $t$-test of the difference of the linear regression coefficients).
Positive connectivity decayed with distance ($p<10^{-6}$ in each of the five sites for positive interactions and $p<0.05$ for negative interactions, $t$-test on the logistic regression coefficient) (Fig.~\ref{fig:6} E), so that cells separated laterally by less than 25 $\mu\mbox{m}$ were 3.2 times more likely to be connected than cells separated laterally by more than 150 $\mu\mbox{m}$. Although the positive connectivity appeared to decay faster with vertical than with lateral distance, the differences in slopes of the respective logistic regression models were not significant with available data. The negative connectivity decayed slower with distance (Fig.~\ref{fig:6} E and F): The slope in the respective logistic models with respect to the lateral distance was significantly higher for positive than for negative connectivities ($p<0.05$ in each of the five sites, two-sample $t$-test of the difference of the logistic regression coefficients).

@@

Moreover, this estimation approach provides a graphical representation of the dependencies in the data that can be used to formulate and test hypotheses about the structure of connectivity in the microcircuit. Importantly, the inferred functional interactions were substantially different from the network of the highest sample correlations.  For example, the $C_{\sf sparse+latent}$ estimator reveals a large number of negative interactions that were not present in the sample correlation matrix (Fig.~\ref{fig:4} F) and may reflect inhibitory circuitry.

Distances between cells in physical space and in sensory feature space had a stronger effect on the partial correlations estimated by the $C_{\sf sparse+latent}$ estimator than on sample correlations (Fig.~\ref{fig:6} A--C).
These differences support the idea that correlations are built up from partial correlations in chains of intermediate cells positioned closer and tuned more similarly to one another, with potentially closer correspondence to anatomical connectivity.  These differences may also be at least partially explained by a trivial effect of regularization: the $L_1$ penalty applied by the estimator (Eq.~\ref{eq:ma}) suppresses small partial correlations to greater extent than large partial correlations, enhancing the apparent effect of distance and tuning.  Distances between cells in physical space and in sensory feature space also had distinct effects on the positive and negative connectivities inferred by the estimator (Fig.~\ref{fig:6} D--F); we could not think of a numerical artifact that could produce this effect. The distinct positive and negative connectivity patterns  may reflect geometric and graphical features of local excitatory and inhibitory networks. Indeed, the relationships between patterns of positive and negative connectivities inferred by the estimator mirrored the properties of excitatory and inhibitory synaptic connectivities with respect to distance, cortical layers, and feature tuning \cite{Song:2005, Oswald:2008, Adesnik:2010, Perin:2011, Fino:2011, Hofer:2011, Isaacson:2011, Levy:2012}. For example, while excitatory neurons form synapses within highly specific local cliques \cite{Perin:2011}, inhibitory interneurons form synapses with nearly all excitatory cells within local microcircuits \cite{Fino:2011, Hofer:2011, Packer:2011}.  To further investigate the link between synaptic connectivity and inferred functional connectivity, in future experiments, we will use molecular markers for various cell types with follow-up multiple whole-cell \emph{in vitro} recordings \cite{Hofer:2011, Ko:2013} to directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological functions. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections).

@@

The measured fluorescent traces were deconvolved to reconstruct the firing rates for each neuron: First, the first principal component was subtracted from the raw traces in order to reduce common mode noise related to small cardiovascular movements \cite{Cotton:2013}. The resulting traces were high-pass filtered above 0.1 Hz and downsampled to 20 Hz (Fig.~\ref{fig:2} C). Then, the firing rates were estimated using by nonnegative deconvolution \cite{Vogelstein:2010}.

@@

Since each of the regularized estimators had one or two hyperparameters, we used \emph{nested cross-validation}:  The outer loop evaluated the performance of the estimators with the hyperparameter values optimized by cross-validation within the inner loop.  Hyperparameters were optimized by a two-phase search algorithm: random search to find a good starting point for the subsequent pattern search to find the global minimum.  The inner cross-validation loop subdivided the training dataset from the outer loop to perform 10-fold cross-validation in order to evaluate each choice of the hyperparameter values.  Thus the size of the training dataset within the inner loop comprised 81\% of the entire recording. Fig.~S2 illustrates the dependence of the validation loss on the hyperparameters of the $C_{\sf sparse+latent}$ estimator for the example site shown in Figures \ref{fig:2} and \ref{fig:4} and the optimal value found by the pattern search algorithm.

>>>>

%% Include all macros below. Please limit the use of macros.
\DeclareMathOperator{\Tr}{tr}
\newcommand{\loss}[1]{\mathcal L(#1)}
\newcommand{\T}{{\sf T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\DeclareMathOperator*{\argmin}{arg\,min}
%% END MACROS SECTION

@@

The evaluation of any covariance matrix estimator, $C$, is performed with respect to a \emph{loss function} $\ell(C,\Sigma)$ to quantify its discrepancy from the truth, $\Sigma$.  The loss function is chosen to attain its minimum when $C=\Sigma$.
Here, in the role of the loss function we adopted the Kullback-Leibler divergence between multivariate normal distributions with equal means, scaled by $\frac 1 p$ to make its values comparable across different population sizes: 
\begin{equation}\label{eq:loss}
    \ell(C,\Sigma) = 
    \frac 1 p D_{KL}\left(\mathcal N_\Sigma\,\Vert\,\mathcal N_C\right) = 
    \frac 1 {2p} \left[\Tr(C^{-1}\Sigma) + \ln\det C - \ln\det\Sigma - p\right] 
\end{equation}
Therefore, $\ell(C,\Sigma)$ is expressed in nats/neuron per time bin.

We drew 30 independent samples with sample sizes $n=250$, 500, 1000, 2000, and 4000 from each model and computed the loss $\ell(C,\Sigma)$ for each of the five estimators.  
The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  
All the regularized estimators produced better estimates (lower loss) than the sample covariance matrix.  
However, estimators whose structure matched the true model outperformed the other estimators (Fig.~\ref{fig:1} Row 5).

Note that when the ground truth had zero correlations (Column A), $C_{\sf factor}$ performed nearly equally well to $C_{\sf diag}$ because it correctly identified zero factors and only estimated the individual variances. 
Similarly, when the number of latent units was zero (Column C), $C_{\sf sparse+latent}$ performed equally well to $C_{\sf sparse}$ because it correctly inferred zero latent units.
With increasing sample sizes, all estimators converged to the truth (zero loss).  
However, the discriminability between their performances only increased with sample size (data not shown).

In more realistic conditions, when the ground truth is not accessible, the loss cannot be computed directly but may be estimated from data through \emph{validation}.
In a validation procedure, a validation sample covariance matrix $C_{\sf sample}^\prime$ is computed from a testing data set that is independent from the data used for computing $C$.
Then the \emph{validation loss} function $\loss{C,C^\prime_{\sf sample}}$ measures the discrepancy of $C$ from $C^\prime_{\sf sample}$. 
Here, in the role of validation loss, we adopted the negative multivariate normal log likelihood of $C$ given $C^\prime_{\sf sample}$, also scaled by $\frac 1 p$ and omitting the constant term: 
\begin{equation}\label{eq:vloss}
    \loss{C,C^\prime_{\sf sample}} = \frac 1 {2p} \left[\Tr(C^{-1}C^\prime_{\sf sample})+\ln\det C \right]
\end{equation}

Since $\loss{\cdot,\cdot}$ is additive in its second argument and $C^\prime_{\sf sample}$ is an unbiased estimate of $\Sigma$, then, for given $C$ and $\Sigma$, validation loss is an unbiased estimate of true loss up to a constant:
\begin{equation}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma} = \ell(C,\Sigma) + \mbox{const}.
\end{equation}
Therefore, the validation procedure allows comparing the relative values of loss of different covariance estimators.

Indeed, validation loss computed by 10-fold cross-validation (see Methods) accurately reproduces the relative values of the loss and the rankings of the covariance estimators without access to the ground truth (Fig.~\ref{fig:1} Row 6).

In the example above, the data were sampled from  multivariate normal random variables. In such models, partial correlations perfectly characterize the conditional dependencies between variables and the graphical models of partial correlations exactly correspond to the conditional dependencies in the data. 
To demonstrate that estimator rankings were robust to deviations from Gaussian models, as long as dependencies between variables included a strong linear component, we repeated the same cross-validated evaluation using pairwise Ising models to generate the data.

Ising models have been used to infer functional connectivity from neuronal spike trains \cite{Hertz:2011}. 
Conveniently, the Ising model has equivalent mathematical form to the Gaussian distribution,
\begin{equation}\label{eq:ising}
    x\sim \frac 1 {Z(J,h)} \exp\left(\frac 1 2 x^\T J x + h^\T x \right)
\end{equation}
but the Ising model is defined on the multivariate binary domain rather than the continuous domain. 
Both models are maximum-entropy models constrained match the mean firing rates and the covariance matrix \cite{Jaynes:1957}.
The partition function $Z(J,h)$ normalizes the distribution on the models' respective domains. 
In the Gaussian model, the matrix $-J^{-1}$ is the covariance matrix; and the mean values are $\mu=J^{-1}h$.  
For the Ising model, $J$ is the matrix of pairwise interactions and $h$ is the vector of the cells' individual activity drives although they do not have a simple relationship to the means and the covariance matrix. 
Both distributions have the same structure of pairwise conditional dependencies: zeros in the matrix $J$ indicate conditional independence between the corresponding pair of neurons.

Indeed, despite their considerable departure from strictly linear conditional dependencies, Ising models yielded the same relationships between the performances covariance estimators as the Gaussian models in cross-validation (Fig.~\ref{fig:2}). Identical interaction matrices $J$  of the joint distributions over the observable and latent variables were used for both the Gaussian and the Ising models.

This simulation study demonstrated that cross-validated evaluation of regularized estimators of the covariance matrix of population activity can discriminate between structures of dependencies in the population. The identity of the most efficient covariance estimator for a particular neural circuit is therefore an empirical finding characterizing the nature of circuit  interactions.

\subsection*{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (Fig.~\ref{fig:3} A--B) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (Fig.~\ref{fig:3} C and D).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.~\ref{fig:3} C). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (Fig.~\ref{fig:3} D). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (Fig.~\ref{fig:3} E and F). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (Fig.~\ref{fig:6} D).

@@

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Fig.~S1 and  Methods). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:4}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<0.01$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

\subsection*{Structure of $C_{\sf sparse+latent}$ estimates}
We examined the composition of the $C_{\sf sparse+latent}$ estimates at each imaged site (Fig.~\ref{fig:5} and Fig.~\ref{fig:6}). Although the regularized estimates were similar to the sample correlation matrix (Fig.~\ref{fig:5} A and B), the corresponding partial correlation matrices differed substantially (Fig.~\ref{fig:5} C and D). The estimates separated two sources of correlations: a network of linear interactions expressed by the sparse component of the inverse and latent units expressed by the low-rank components of the inverse (Fig.~\ref{fig:5} E). The sparse partial correlations revealed a network that differed substantially from the network composed of the greatest coefficients in the sample correlation matrix (Fig.~\ref{fig:5} F, G, H, and I).

In the example site (Fig.~\ref{fig:5}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $\mbox{connectivity}=1-\mbox{sparsity}$) with average node degree of 20.9 (Fig.~\ref{fig:5} G). The average node degree, \emph{i.e.}\;the average number of interactions linking each neuron, is related to connectivity as $\mbox{degree} = \mbox{connectivity}\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units increased with population size (Fig.~\ref{fig:6} A) but the connectivity was highly variable (Fig.~\ref{fig:6} B): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and warrants further investigation.

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 27 sites were about 5 times lower than the average sample correlations (Fig.~\ref{fig:6} C). This suggests that correlations between neurons build up from multiple chains of smaller interactions. Furthermore, the average partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.45 whereas that of the average partial correlations was 0.29, with larger populations exhibiting greater uniformity of average partial correlations than the smaller populations.

While the sample correlations were mostly positive, the sparse component of the partial correlations (`interactions') had a high fraction (28.7\% in the example site) of negative values (Fig.~\ref{fig:5} F). The fraction of negative interactions increased with the inferred connectivity (Fig.~\ref{fig:6} D), suggesting that negative interactions can be inferred only after a sufficient density of positive interactions has been uncovered.

Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013, Sadovsky:2014}.  We therefore compared the interactions in the sparse component of $C_{\sf sparse+latent}$ to those obtained from the sample correlations thresholded to the same level of connectivity. The networks revealed by the two methods differed substantially. In the example site with 7.2\% connectivity in $C_{\sf sparse+latent}$, only 27.7\% of the connections coincided with the above-threshold sample correlations (Fig.~\ref{fig:5} F, H, and I). In particular, most of the inferred negative interactions corresponded to low sample correlations (Fig.~\ref{fig:5} F) where high correlations should be expected given the rest of the correlation matrix.

@@

We examined how the structure of the $C_{\sf sparse+latent}$ estimates related to the differences in orientation preference and to the physical distances separating pairs of cells (Fig.\;\ref{fig:7}).  Five sites with highest pairwise connectivities were included in the analysis. Partial correlations were computed using Eq.~\ref{eq:partial} based on the regularized estimate, including both the sparse and the latent component. Connectivity was computed as the fraction of pairs of cells connected by non-zero elements (interactions) in the sparse component of the estimate, distinguishing between the positive and negative connectivities.

First, we analyzed how correlations and connectivity depended on the difference in preferred orientations ($\Delta \mbox{ori}$) of pairs of significantly ($\alpha=0.05$) tuned cells. The partial correlations decayed more rapidly with $\Delta\mbox{ori}$ than did sample correlations ($p<10^{-9}$ in each of the five sites, two-sample $t$-test of the difference of the linear regression coefficients). Positive connectivity decreased with $\Delta\mbox{ori}$ ($p<0.005$ in each of the five sites, $t$-test on the logistic regression coefficient) whereas negative connectivity did not decrease (Fig.~\ref{fig:7} D): The slope in the logistic model of connectivity with respect to $\Delta\mbox{ori}$ was significantly higher for positive than for negative interactions ($p<0.04$ in each of the five sites, two-sample $t$-test of the difference of the logistic regression coefficient).

Second, we compared how correlations and connectivity depended on the physical distance separating pairs of cells. We distinguished between lateral distance, $\Delta x$, in the plane parallel to the pia, and vertical distance, $\Delta z$, orthogonal to the pia.  When considering the dependence on $\Delta x$, the analysis was limited to cell pairs located at the same depth with $\Delta z < 30\,\mu\mbox{m}$; conversely, when considering the dependence on $\Delta z$, only vertically aligned cell pairs with $\Delta x < 30\,\mu\mbox{m}$ were included. Again, the partial correlations decayed more rapidly both laterally and vertically than sample correlations ($p<10^{-6}$ in each of the five sites, for both lateral and vertical distances, two-sample $t$-test of the difference of the linear regression coefficients).
Positive connectivity decayed with distance ($p<10^{-6}$ in each of the five sites for positive interactions and $p<0.05$ for negative interactions, $t$-test on the logistic regression coefficient) (Fig.~\ref{fig:7} E), so that cells separated laterally by less than 25 $\mu\mbox{m}$ were 3.2 times more likely to be connected than cells separated laterally by more than 150 $\mu\mbox{m}$. Although the positive connectivity appeared to decay faster with vertical than with lateral distance, the differences in slopes of the respective logistic regression models were not significant with available data. The negative connectivity decayed slower with distance (Fig.~\ref{fig:7} E and F): The slope in the respective logistic models with respect to the lateral distance was significantly higher for positive than for negative connectivities ($p<0.05$ in each of the five sites, two-sample $t$-test of the difference of the logistic regression coefficients).

@@

Moreover, this estimation approach provides a graphical representation of the dependencies in the data that can be used to formulate and test hypotheses about the structure of connectivity in the microcircuit. Importantly, the inferred functional interactions were substantially different from the network of the highest sample correlations.  For example, the $C_{\sf sparse+latent}$ estimator reveals a large number of negative interactions that were not present in the sample correlation matrix (Fig.~\ref{fig:5} F) and may reflect inhibitory circuitry.

Distances between cells in physical space and in sensory feature space had a stronger effect on the partial correlations estimated by the $C_{\sf sparse+latent}$ estimator than on sample correlations (Fig.~\ref{fig:7} A--C).
These differences support the idea that correlations are built up from partial correlations in chains of intermediate cells positioned closer and tuned more similarly to one another, with potentially closer correspondence to anatomical connectivity.  These differences may also be at least partially explained by a trivial effect of regularization: the $L_1$ penalty applied by the estimator (Eq.~\ref{eq:ma}) suppresses small partial correlations to greater extent than large partial correlations, enhancing the apparent effect of distance and tuning.  Distances between cells in physical space and in sensory feature space also had distinct effects on the positive and negative connectivities inferred by the estimator (Fig.~\ref{fig:7} D--F); we could not think of a numerical artifact that could produce this effect. The distinct positive and negative connectivity patterns  may reflect geometric and graphical features of local excitatory and inhibitory networks. Indeed, the relationships between patterns of positive and negative connectivities inferred by the estimator mirrored the properties of excitatory and inhibitory synaptic connectivities with respect to distance, cortical layers, and feature tuning \cite{Song:2005, Oswald:2008, Adesnik:2010, Perin:2011, Fino:2011, Hofer:2011, Isaacson:2011, Levy:2012}. For example, while excitatory neurons form synapses within highly specific local cliques \cite{Perin:2011}, inhibitory interneurons form synapses with nearly all excitatory cells within local microcircuits \cite{Fino:2011, Hofer:2011, Packer:2011}.  To further investigate the link between synaptic connectivity and inferred functional connectivity, in future experiments, we will use molecular markers for various cell types with follow-up multiple whole-cell \emph{in vitro} recordings \cite{Hofer:2011, Ko:2013} to directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological functions. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections).

@@

The measured fluorescent traces were deconvolved to reconstruct the firing rates for each neuron: First, the first principal component was subtracted from the raw traces in order to reduce common mode noise related to small cardiovascular movements \cite{Cotton:2013}. The resulting traces were high-pass filtered above 0.1 Hz and downsampled to 20 Hz (Fig.~\ref{fig:3} C). Then, the firing rates were estimated using by nonnegative deconvolution \cite{Vogelstein:2010}.

@@

Since each of the regularized estimators had one or two hyperparameters, we used \emph{nested cross-validation}:  The outer loop evaluated the performance of the estimators with the hyperparameter values optimized by cross-validation within the inner loop.  Hyperparameters were optimized by a two-phase search algorithm: random search to find a good starting point for the subsequent pattern search to find the global minimum.  The inner cross-validation loop subdivided the training dataset from the outer loop to perform 10-fold cross-validation in order to evaluate each choice of the hyperparameter values.  Thus the size of the training dataset within the inner loop comprised 81\% of the entire recording. Fig.~S1 illustrates the dependence of the validation loss on the hyperparameters of the $C_{\sf sparse+latent}$ estimator for the example site shown in Figures \ref{fig:3} and \ref{fig:5} and the optimal value found by the pattern search algorithm.
